---
title: "Detroit Metro Land Cover Change Modeled by Cellular Automata"
author: "K. Arthur Endsley"
date: 'Sunday, November 23, 2014'
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r preamble, warning=FALSE, message=FALSE, results=FALSE}
library(sp, raster)
library(rgdal)
library(plyr, reshape2)
library(bnlearn)

setwd('/usr/local/dev/rdetroit/')
options(stringsAsFactors=FALSE)
```

# Preparation

## Land Cover Data

We'll try aggregating the data to 90 meters to see if that gives us a feasible grid with which to work.

```{r}
file.loc <- '/home/arthur/Workspace/TermProject/'

require(raster)
rast2000 <- raster::raster(paste0(file.loc, 'nlcd2001_nad83.tif'))
rast2006 <- raster::raster(paste0(file.loc, 'nlcd2006_nad83.tif'))
reclass.matrix <- matrix(c(c(0,10,0), c(10,11,NA), c(12,20,0),
                           c(20,23,1), c(23,24,2), c(24,99,0)),
                         byrow=TRUE, ncol=3)
dev2001 <- raster::reclassify(rast2000, reclass.matrix, right=TRUE) # Intervals closed on right
dev2006 <- raster::reclassify(rast2006, reclass.matrix, right=TRUE)
```

```{r}
plot(dev2001); title('2001 Development Intensity at 30 meters')
```

```{r}
counties <- readOGR(paste0(file.loc, 'ancillary/co26_d00_select_nad83.shp'),
                    'co26_d00_select_nad83')
counties <- spTransform(counties, crs(dev2000))

plot(counties, col='#EEEEEE')
plot(dev2006 - dev2001, breaks=c(-2, -1, 0, 2), add=TRUE,
     col=c('#77DD00', '#FFFFFF00', '#113355'))
title('Change in Metro Detroit Development, 2001-2006')
```

We'll first aggregate the land cover data to 300 meters.

```{r}
dev2001 <- aggregate(dev2001, fact=10, fun=modal)
dev2006 <- aggregate(dev2006, fact=10, fun=modal)
save(dev2001, dev2006, file='rda/caAggregates.rda')
```

```{r}
plot(dev2001); title('2001 Development Intensity at 300 meters')
```

## Landscape and Census Data

We'll grab the other spatial data we created when training the Bayesian network.

```{r}
file.loc <- '/home/arthur/Workspace/TermProject/'
load(file='rda/spatialMeasures.rda')
load(file='rda/caAggregates.rda') # Replace dev2001 and dev2006 with aggregated

# Recreation and outdoor areas
rec.area.dist <- raster::raster(paste0(file.loc,
                                       'ancillary/rec+outdoor_nad83_prox_cut.tiff'))
rec.area.dist <- resample(rec.area.dist, dev2001)
```

## Rasterizing

```{r}
vars <- c('married.hholds', 'med.hhold.income', 'pop.density', 'male.pop')
layers <- as.list(1:length(vars))
names(layers) <- vars
layers$old <- as.factor(dev2001)
layers$rec.area.proximity <- rec.area.dist

for (var in vars) {
  layers[var] <- rasterize(attr2006, dev2006, var)
}

save(layers, file='rda/layers.rda')
load(file='rda/layers.rda')
```

## Masking and Discretizing

First, we want to mask out the pixels outside of the census tracts.

```{r}
layers$old <- mask(dev2001, layers$pop.density, maskvalue=NA)
layers$rec.area.proximity <- mask(layers$rec.area.proximity,
                                  layers$pop.density, maskvalue=NA)
```

Next, we must discretize all the census and landscape layers.

```{r}
load(file='rda/graphs.rda')

# Create a reclass matrix from the levels
vars <- c('rec.area.proximity', 'married.hholds', 'med.hhold.income', 'pop.density', 'male.pop')
layer.levels <- list(rec.area.proximity=levels(training.discrete$rec.area.proximity),
                     married.hholds=levels(training.discrete$married.hholds),
                     med.hhold.income=levels(training.discrete$med.hhold.income),
                     pop.density=levels(training.discrete$pop.density),
                     male.pop=levels(training.discrete$male.pop))

for (var in vars) {
  # Split apart e.g. "(20.1,190]"
  reclass.matrix <- sapply(levels(training.discrete[,var]),
                           function (s) as.numeric(unlist(strsplit(gsub('\\[|\\]|\\(|\\)',
                                                                        '', s), ','))))
  reclass.matrix[1] <- 0
  layers[var] <- raster::reclassify(get(var, layers),
                                    cbind(reclass.matrix, c(0, 1)),
                                    right=TRUE) # Intervals closed on right
}
```

# Stacking

```{r}
layers <- stack(layers)

save(layers, file='rda/layerStack.rda')
load(file='rda/layerStack.rda')

# Clean-up
remove(var, vars, rec.area.dist, roads.dist, reclass.matrix)
```

# Examining the Network

```{r message=FALSE, warning=FALSE}
load(file='rda/graphs.rda')

require(gRain)

# We use the junction tree algorithm to create an independence network that we can query
jtree <- compile(as.grain(fit.expert))

vars <- c('married.hholds', 'med.hhold.income', 'pop.density', 'male.pop', 'old')
foo <- setEvidence(jtree, nodes=vars, nslist=as.list(na.omit(getValues(layers, 1, 1))[1,]))
```



