---
title: Studying Urban Change in Detroit by Modeling Impervious Surface Changes with Bayesian Networks
author: "K. Arthur Endsley"
date: "December 9, 2014"
output:
  pdf_document:
    fig_caption: true
bibliography: "~/Mendeley.bib"
csl: TermPaper.csl
smart: true
fontsize: 12pt
geometry: margin=1in
documentclass: report
classoption: notitlepage
---

# Background

There are many motivations for studying urban environments and urban change in terms of land use or land cover changes. Examples of studies in this area include mitigating the impacts of urban sprawl [@Jantz2004], generating urban development scenarios [@Kocabas2007], or monitoring rates of urban growth and impervious surface increase [@Sexton2013]. To these ends, previous studies have employed a variety of models that attempt either to predict future future states of the landscape or to explain the drivers of urban change. Any land cover model can be located on this continuum between \textit{projection} and \textit{explanation} [@Brown2013a] and models generally excel at one purpose but are weak or not even applicable for the other.

## Case Study: Detroit

# Methods

This study employs a land cover model driven by spatially-explicit socioeconomic, demographic, and landscape variables (e.g., distance to roads). Predictions of future land cover are made based on these variables and the previous or currently-observed land cover. Specifically, \textit{development intensity} as indicated by the proportion of impervious surface area is predicted for the Detroit metropolitan area for two years, 2006 and 2011.

The methods are described in detail in the following sections but are here broadly summarized. Land cover data, U.S. Census data, and derived landscape measures were spatially joined, randomly sampled, and then used to train the Bayesian network, a process that includes determining the network structure and then fitting the parameters. In prediction, the predictors (land cover, census, and landscape measures) are first aggregated to 300 meters, ten times the finest resolution of the available data. Estimations of future land cover are therefore also at a scale of 300 meters. All of the spatially explicit predictor datasets, available as geographic information systems (GIS) layers, were projected to a common coordinate reference system and clipped to the study area.

The analysis was primarily carried out within the R statistical computing environment [@R2014] using a number of different packages made freely available to the scientific community including \texttt{sp}, \texttt{raster}, \texttt{rgdal}, \texttt{plyr}, and \texttt{reshape2}. Visualizations were aided by the perceptually linear and colorblind-friendly palettes available in the \texttt{RColorBrewer} package and based on Cynthia Brewer's innovative work in color theory [@Brewer2014]. The structure learning, parameter fitting, and operation of Bayesian networks were facilitated by the \texttt{bnlearn} package [@Scutari2014]. Cohen's kappa, a measure of inter-rater reliability and agreement, was calculated using the \texttt{irr} package.

## Data Sources

### Land Cover Data

The land cover data used in this study comes from the National Land Cover Dataset (NLCD) and was obtained from the U.S. Geological Survey. Land cover data in this dataset are available at 30-meter resolution every five years from 2001 to 2011 (thus far). The NLCD classes are consistent between the 2001, 2006, and 2011 land cover maps and the the Multi-Resolution Land Characteristics Consortium (MRLC) asserts that they can be used for longitudinal studies from 2001 to 2011 [@USGS2014]. The classes are several and most are not relevant to this study. Consequently, the NLCD data for 2001, 2006, and 2011 were reclassified to encompass only three classes (Figure 1). These three classes represent undeveloped areas, areas with less than 80% impervious surface cover, and areas that are 80% or more covered by impervious surface.

```{r reclass.explication, message=FALSE, warning=FALSE, echo=FALSE}
require(pander)
panderOptions('table.split.table', Inf)
pander::set.caption('Reclassification table for the National Land Cover Dataset (NLCD)')
pander(data.frame('Input Class Range'=c('[0-10] (Unused in NLCD)',
                  '11 ("Open Water")', '[12-20] ("Ice and Snow")',
                  '[21-23] ("Open Space, Low and Medium-Intensity Development")',
                  '24 ("High-Intensity Development")',
                  '[25-99] (Vegetation, Cultivation and Wetlands)'),
                  'Output Class'=c(0,NA,0,1,2,0),
                  'New Class Label'=c('Undeveloped', 'Excluded', 'Undeveloped',
                                      'Low-Intensity Development',
                                      'High-Intensity Development', 'Undeveloped'),
                  check.names=FALSE),
       justify=c('left', 'center', 'left'),
       style='grid')
```

### Socioeconomic and Demographic Data

U.S. Census and American Community Survey (ACS) data were used to obtain socioeconomic (e.g., median household income) and demographic (e.g., sex and racial make-up) data on Detroit communities. Ordinarily, obtaining the comprehensive census measures desired in an open-ended study directly from a U.S. government website can be difficult. To expedite data processing, these U.S. Census and ACS data were downloaded from SocialExplorer.org [@Explorer2014], a website that offers a user-friendly interface for downloading census data by census tract or other geographic subset.

As census tract boundaries can change between census years (and did change between 2001 and 2010), a necessary first step to using the census data in this study included cross-walking 2001 census data to the 2010 census tract boundaries. This was facilitated by the use of crosswalk tables developed by Logan et al. [-@Logan2014]. Census data were also normalized so as to eliminate the effects of census tract land area, total population, total number of housing units, and total household population. Some census variables, such as population density and median household income, were intrinsically scale-independent and did not need to be normalized. The census variables were chosen so that only variables jointly present across the 2001 U.S. Census, 2006-2010 ACS, and 2008-2012 ACS were considered. It is worth remarking that while this ensures every training and prediction dataset has the same variables, the model used in this paper does not require simultaneous observations of every variable (that is, it tolerates missing data).

### Landscape Measures

Two additional predictors were initially considered: distance to major roads and distance to parks and open recreation areas. These were derived from maps of roads, obtained from the Michigan Geographic Data Library [@Michigan2014], and of parks and open recreation areas, obtained from the Southeastern Michigan Council of Governments, and were both rasterized to 30 meters, matching the resolution of the land cover data. Proximities to roads and parks were then calculated from these layers using the Geospatial Data Abstraction Library [@Team2014].

## Finding Independent Predictors

After the reclassified land cover data, census measures, and landscape data were spatially joined in R, Pearson's correlation coefficients were calculated for every pair of independent predictors. A threshold of $R^2=0.5$ was chosen as the cut-off for collinearity; a correlation above this threshold was interpreted to mean that one of the two variables had to be left out of the model. Although multicollinearity is not a serious concern for purely predictive models, as the objective of this model is explanatory, minimizing multicollinearity is essential so as to avoid artificially inflating parameter values.

A number of highly correlated "clusters" were located from these pair-wise tests. Population density, number of married households, and number of family households were one such cluster. Median household income, proportion of housing occupied, proportion of housing occupied by the owner, and proportion of the population poor or struggling were found to be correlated with one another. Male and female populations were obviously highly correlated with one another. Other census measures, including any measure of racial make-up, were all highly correlated with one another. From these clusters, independent measures were somewhat arbitrarily chosen. For instance, male population was selected where the variable female population would have presumably been equivalent in the model. Population density and median household income, however, were chosen specifically because it is assumed that such data are more likely to be available on fine time scales from sources other than the U.S. Census.

Distance to roads and population density were found to be marginally correlated ($R^2\approx 0.5$) with previous ("old") land cover and therefore were dropped from the final model (as dropping "old" land cover is not an option). The final variables included in the model were therefore: Median household income, proportion of the population poor or struggling, proportion of the population that is male, old and new land cover, and proximity to parks and open areas. All the predictor variables were discretized into two quantiles (a median split of the data) with the exception of previous land cover, which, as with future land cover, was already discretized into three classes.

## The Model

The modeling approach in this study involves the use of Bayesian networks, which are also known variously as belief networks, Bayesian belief networks, Bayes nets, and causal probabilistic networks [@Uusitalo2007]. Bayesian networks are so-named for the Reverend Thomas Bayes (1702-1760) who is credited with first positing what has become known as Bayes' theory, a statement about conditional probabilities that allows for the calculation of "inverse probabilities" or the probability of one of a cause given its consequences.

Bayesian networks (BNs) are directed, acyclic graphs [@Nagarajan2013]. Each node in the graph is a variable and no meaningful discting between "predictor" or "response" variable is made. Nodes connected to other nodes imply a conditional dependence in a certain direction (hence, the graph is directed) and the links between them cannot form cycles (hence, the graph is acyclic). BNs must also exhibit the Markov property; that is, the conditional probability of any node must depend only on its immediate parents. A more rigorous statement of this property is that the nodes must be \textit{d-separated}, a concept that is beyond the scope of this paper (see Charniak et al. [-@Charniak1991] and Nagarajan et al. [-@Nagarajan2013] for more details).

In general, BNs are either discrete or continuous; discrete and continuous variables are usually not mixed and software tools that support mixed types in the network are not common [@Uusitalo2007;@Nagarajan2013]. This is to facilitate calculating the joint probability distribution, which is either a multinomial distribution in the case of discrete-valued variables or a Gaussian distribution in the case of continuous values. When the variables that make up the nodes take on continuous values, the probability distribution of each node is a normal distribution and the resulting network that is formed is termed a Gaussian Bayesian network. In the case of continuous variables, the parameters are just regression coefficients. Because the nodes of a Bayesian network are linked, multivariate regression is performed to predict the distribution arising at each node in the network, providing regression coefficients for each pairwise interaction between a node and its connections [@Nagarajan2013].

Training a Bayesian network generally consists of two steps: learning the network structure and then fitting the parameters. In some studies, the network structure may be known or specified by an expert. The conditional probability tables (CPTs) for some or all of the variables might also be specified by an expert [@Kocabas2007]. In this study, there was not sufficient domain knowledge on hand to esimate the CPTs but an "expert" network structure was specified

### Learning Network Structure

Structure learning is computationally intensive but many different algorithms are available that are all tractable on end-user hardware. Three classes of network learning---algorithms, constraint-based, score-based, and hybrid algorithms---were used to investigate possible network structures and to select a stable (consistent) structure for modeling. The algorithms used were all available in the \texttt{bnlearn} package for R. A random sample of the discrete training data was used to learn the network structure using a variety of constraint-based, score-based, and hybrid network learning algorithms. Most of the algorithms produced extremely dense graphs including many complete graphs. Ultimately, two hybrid algorithms, General 2-Phase Restricted Maximization (RSMAX2) and Max-Min Hill Climbing (MMHC), agreed upon the same network structure. As this class of algorithms is considered to produce more reliable networks than alternative algorithms [@Nagarajan2013], their network structure was used for the "learned" model.

In addition to the learned network, an "expert" network was specified based on the author's beliefs about conditional dependencies between predictor variables. When scored using a second random sample disjoint from the fitting sample, the expert network was found to be superior to the learned network according to Bayes' Information Criterion (BIC). This may only be because the expert network is more sparse than the learned network.

### Fitting the Parameters

Parameter learning is generally done through a maximum likelihood approach (whereby the best fit parameters are estimated) or a Bayesian approach (whereby the posterior distribution of the parameters for a discrete distribution is estimated). The Bayesian approach is preferred as it provides more robust estimates and guarantees the conditional probability tables will be complete [@Nagarajan2013]. This is the approach used in this paper.

The BNs were trained from the high-resolution, 30-meter land cover data and landscape measures joined to the coarse-resolution census data. The random samples used to learn the network structure, score the network structure, and fit the model were distinct; a three-folds random sample was created for these purposes. Each disjoint sample contains less than 4% of the complete dataset. The "expert" and the "learned" networks were fitted with the same sample.

## Prediction

The predictor variables were aggregated to 300 meters to reduce the computational complexity. Prediction with the either the expert or learned Bayesian network consists of the same, basic steps:

\begin{enumerate}
\item{For each pixel, get the available evidence (census measures, proximities, and land cover observations).}
\item{Obtain the posterior probability distributions given the evidence.}
\item{Considering the ``new'' land cover variable, choose the outcome (land cover) from the posterior probability distribution.}
\end{enumerate}

One of the virtues of BNs is that predictions do not require a simultaneous observation of all predictor variables; even just one predictor variable can be used as evidence. This would be useful, for instance, in the prediction of 2011 land cover prediction from the 2010 U.S. Census as this census did not include the "median household income" or "population poor or struggling" variables, which had been used to develop and train the network. The predictions made before evidence and after seeing the evidence can be appreciated graphically in Figure 2, which shows an arbitrary prediction made one of the BNs before the "no data" areas were masked.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=5, fit.height=4, fig.cap='A Bayesian network prediction without masking'}
require(raster)
require(rgdal)
load(file='rda/outputsUnmasked2011.rda')

require(RColorBrewer)
cols <- brewer.pal(3, 'YlGnBu')

par(xpd=FALSE)
plot(output.expert.2011, col=cols, axes=FALSE, box=FALSE, legend=FALSE,
     xlab='Easting (meters)', ylab='Northing (meters)')
axis(1); axis(2);
par(xpd=TRUE)
legend('right', bty="n", title='', fill=cols, inset=c(-0.4, 0),
       legend=c('Undeveloped', 'Low-Intensity', 'High-Intensity'))
```

# Results

## Model Criticism

# Conclusions

## Suggested Improvements

# References

\raggedright